---
title: "ESM 262 - Database Assignment"
author: "Kenneth Larrieu"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(DBI)
knitr::opts_chunk$set(echo = TRUE)
```

## Import and Tidy

1. Read the gazetteer data as-is (all columns; no type conversion) into a `gaz_raw` tibble.

``` {r}
gaz_raw = read_delim("CA_Features_20180401.txt", "|", col_types = cols(.default = col_character()))
gaz_raw
```

2. Select columns to include in `gaz` tibble.
``` {r}
gaz = gaz_raw %>% select(FEATURE_ID,
                          FEATURE_NAME,
                          FEATURE_CLASS,
                          STATE_ALPHA,
                          COUNTY_NAME,
                          PRIM_LAT_DEC,
                          PRIM_LONG_DEC,
                          SOURCE_LAT_DEC,
                          SOURCE_LONG_DEC,
                          ELEV_IN_M,
                          MAP_NAME,
                          DATE_CREATED,
                          DATE_EDITED)
gaz
```
3. Convert `gaz` columns to correct type. Convert any placeholders for unknown data to `NA`
```{r}
gaz = gaz %>%
  type_convert(col_types = cols(
    FEATURE_ID = col_integer(),
    PRIM_LAT_DEC = col_double(),
    PRIM_LONG_DEC = col_double(),
    SOURCE_LAT_DEC = col_double(),
    SOURCE_LONG_DEC = col_double(),
    ELEV_IN_M = col_integer(),
    DATE_CREATED = col_date(format = "%M/%d/%Y"),
    DATE_EDITED = col_date(format = "%M/%d/%Y")
          ))

gaz = gaz %>%
  mutate(MAP_NAME = replace(MAP_NAME, MAP_NAME == "Unknown", NA)) %>%
  mutate(PRIM_LAT_DEC = replace(PRIM_LAT_DEC, PRIM_LAT_DEC == 0, NA)) %>%
  mutate(PRIM_LONG_DEC = replace(PRIM_LONG_DEC, PRIM_LONG_DEC == 0, NA)) %>%
  mutate(SOURCE_LAT_DEC = replace(SOURCE_LAT_DEC, SOURCE_LAT_DEC == 0, NA)) %>%
  mutate(SOURCE_LONG_DEC = replace(SOURCE_LONG_DEC, SOURCE_LONG_DEC == 0, NA))
gaz
```


4. Filter out missing primary lat/long, only include places in CA
```{r}
gaz = gaz %>%
  filter(!is.na(PRIM_LAT_DEC) & !is.na(PRIM_LONG_DEC)) %>%
  filter(STATE_ALPHA == "CA")

gaz
```

## Create Database

1. Create a connection to a new ```gaz.db``` SQLite database.

``` {r}
con <- DBI::dbConnect(RSQLite::SQLite(), "gaz.db")

```

2. Copy the ```gaz``` tibble into the ```gaz.db``` database.

``` {r}
con %>% copy_to(gaz, "gaz",
                temporary = FALSE,
                indexes = list(
                  "FEATURE_NAME",
                  "FEATURE_CLASS",
                  "COUNTY_NAME",
                  "SOURCE_LAT_DEC",
                  "SOURCE_LONG_DEC",
                  "PRIM_LAT_DEC",
                  "PRIM_LONG_DEC"
                ))

```


## Analyze

Using SQL queries and the ```dbGetQuery``` function, create R code snippets that answer the following questions about California:

1. What is the most-frequently-occuring feature name?

```{r}
top_feature = con %>% dbGetQuery("
                                  SELECT FEATURE_NAME, max(count) FROM (
                                  SELECT FEATURE_NAME, COUNT(FEATURE_NAME) AS count
                                  FROM gaz
                                  GROUP BY FEATURE_NAME
                                  ORDER BY count desc);
                                 ")
top_feature
```

2. What is the least-frequently-occuring feature class?

```{r}
bottom_class = con %>% dbGetQuery("SELECT FEATURE_CLASS, count FROM (
                                  SELECT FEATURE_CLASS, COUNT(FEATURE_CLASS) AS count
                                  FROM gaz
                                  GROUP BY FEATURE_CLASS
                                  ORDER BY count asc ) WHERE count = 1;
                                  ")
bottom_class
```

3. What is the approximate center point of each county?
```{r}
county_centers = con %>% dbGetQuery("SELECT COUNTY_NAME,
                                    (min(PRIM_LAT_DEC)+max(PRIM_LAT_DEC))/2 AS center_lat_dec,
                                    (min(PRIM_LONG_DEC)+max(PRIM_LONG_DEC))/2 AS center_long_dec
                                    FROM gaz
                                    WHERE (COUNTY_NAME IS NOT NULL) AND
                                    (SOURCE_LAT_DEC IS NULL) AND
                                    (SOURCE_LONG_DEC IS NULL)
                                    GROUP BY COUNTY_NAME
                                    ;")
county_centers

```

4. What are the fractions of the total number of features in each county that are natural? man-made?
Hint: Copy the feature class categories tibble you created into the ```gaz.db``` database. Then you can ```JOIN``` it to the the ```gaz``` table.
```{r}
class_types = gaz %>%
  group_by(FEATURE_CLASS) %>%
  summarize() %>%
  mutate(type = case_when(FEATURE_CLASS %in% c("Airport",
                                            "Bridge",
                                            "Building",
                                            "Canal",
                                            "Cemetery",
                                            "Church",
                                            "Civil",
                                            "Crossing",
                                            "Dam",
                                            "Harbor",
                                            "Hospital",
                                            "Locale",
                                            "Military",
                                            "Mine",
                                            "Oilfield",
                                            "Populated Place",
                                            "Post Office",
                                            "Reservoir",
                                            "School",
                                            "Tower",
                                            "Well")
                          
                                            ~ "manmade",
                          
                          FEATURE_CLASS %in% c("Arch",
                                               "Area",
                                               "Arroyo",
                                               "Bar",
                                               "Basin",
                                               "Bay",
                                               "Beach",
                                               "Bench",
                                               "Bend",
                                               "Cape",
                                               "Cave",
                                               "Channel",
                                               "Cliff",
                                               "Falls",
                                               "Flat",
                                               "Forest",
                                               "Gap",
                                               "Glacier",
                                               "Gut",
                                               "Island",
                                               "Isthmus",
                                               "Lake",
                                               "Lava",
                                               "Pillar",
                                               "Plain",
                                               "Range",
                                               "Rapids",
                                               "Reserve",
                                               "Ridge",
                                               "Sea",
                                               "Slope",
                                               "Spring",
                                               "Stream",
                                               "Summit",
                                               "Swamp",
                                               "Valley",
                                               "Woods")
                          
                                              ~ "natural"
                          ))


con %>% copy_to(class_types, "class_types",
        temporary = FALSE,
        indexes = list("FEATURE_CLASS", "type"))

class_types = con %>% dbGetQuery("SELECT COUNTY_NAME, gaz.FEATURE_CLASS, type
                                    FROM gaz
                                    JOIN class_types
                                    ON gaz.FEATURE_CLASS = class_types.FEATURE_CLASS
                                    WHERE COUNTY_NAME IS NOT NULL
                                    ;")

con %>% copy_to(class_types, "class_types",
        temporary = FALSE,
        indexes = list("COUNTY_NAME", "type"), overwrite = TRUE)


type_counts = con %>% dbGetQuery("SELECT COUNTY_NAME, type, COUNT() as count FROM class_types
                   GROUP BY COUNTY_NAME, type")

con %>% copy_to(type_counts, "type_counts",
                temporary = FALSE,
                indexes = list("COUNTY_NAME", "type", "count"))

totals = con %>% dbGetQuery("
                    SELECT COUNTY_NAME, type, sum(count) AS total FROM type_counts
                    GROUP BY COUNTY_NAME
                   ")

con %>% copy_to(totals, "totals",
                temporary = FALSE,
                indexes = list("COUNTY_NAME", "type", "total"))

fractions = con %>% dbGetQuery("SELECT type_counts.COUNTY_NAME, type_counts.type,
                    round(count*1.0/total, 2) AS fraction
                   FROM type_counts
                   JOIN totals ON type_counts.COUNTY_NAME = totals.COUNTY_NAME")

con %>% copy_to(fractions, "fractions",
                temporary = FALSE, overwrite = TRUE)

manmade_fractions = con %>% dbGetQuery("SELECT COUNTY_NAME, fraction AS manmade_fraction FROM fractions WHERE type = \"manmade\";")

con %>% copy_to(manmade_fractions, "manmade_fractions")

tbl(con,"fractions") %>% show_query()


```

```{r}
con %>% dbDisconnect()
```